---
title: "Tidyverse_3_Dplyr"
author: "J"
date: "25/01/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(readxl)
require(tidyverse)
options(scipen = 999) 
```
#### Exercises: tidyr ####

#### Ensure you have the sexual offences with population as a tidy tibble ####

```{r check sexual offences tidy tibble has added population column}
sexual_offences <- as_tibble(read_excel("sexual_offences_1946_1970.xlsx"))
population <- as_tibble(read_excel("population.xlsx"))
population_data <- population[,2]
sexual_offences_with_population <- sexual_offences  %>%
  add_column(population_data)

sexual_offences_with_pop_tidy <- sexual_offences_with_population %>% 
  gather(offence_name, reports_taken, `abduction-females`:`sex-males`) %>% 
  separate(offence_name, c("offence","against"), sep='-')


```

#### Exercises: dplyr ####

```{r summaries offences by year and victim gender}
offences_against <- sexual_offences_with_pop_tidy %>%
  group_by(year,against)  %>%
  summarise(reports=sum(reports_taken))

```

If you get an output in red warning you about how `summarise` groups objects, you can ignore it for now. Unless it actually says "error" - then you probably need to try something else. 

```{r summarise offences by year and add a column for offences per population}
offences_by_year <- sexual_offences_with_pop_tidy %>%
  group_by(year,population)  %>% 
  summarise(reports=sum(reports_taken)) %>%
  mutate(reports/population)

```
You can, of course, look at the new tibbles that you have created by viewing or printing them... The grammar for both is demonstrated below. 
```{r have a look at your summaries}
view(offences_against)
print(offences_by_year)

```
