---
title: "Tidyverse_4_Merge"
author: "J"
date: "25/01/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(readxl)
require(tidyverse)
```

#### Ensure you have the drugs data sets loaded into R as tidy tibbles ####
```{r check that you have all needed tidy tibbles}
drugs_siezures <- 
  as_tibble(read.csv("drugs_seizures.csv",  header = TRUE, check.names=FALSE, fileEncoding="UTF-8-BOM"))
drugs_volume <- 
  as_tibble(read.csv("drugs_volume.csv", header = TRUE, check.names=FALSE, fileEncoding="UTF-8-BOM"))

drugs_siezures_tidy <- drugs_siezures %>% gather(year,siezures,'2007':'2020') %>%
  separate(drug, c("class","type"), sep='-')

drugs_volume_tidy <- drugs_volume %>% gather(year,volume_siezed,'2007':'2020') %>%
  separate(drug, c("class","type"), sep='-')  %>%
  relocate('unit', .after = 'volume_siezed')

```

#### Exercises: Merging data sets ####

# Start with a full_join
```{r do}
full_join_drugs <- full_join(drugs_siezures_tidy,drugs_volume_tidy)
print(full_join_drugs)

```

#    Explore more with anti-join
```{r find and then summarise the missing volumes.}
missing_volumes <- anti_join(drugs_siezures_tidy, drugs_volume_tidy, by = c("force", "type", "class", "year"))

missing_volumes %>% group_by(type) %>% summarise((n=n()))

```

# Explore other types of joins
Note: you can specify which columns you want to join, as I have done with the inner_join function. But if you do not specify, then R tries to join by everything that can be used to join by, which in this case is `by = c("force", "class", "type", "year")`. It will give you a red message to let you know that it had to do this work for you and that you should know exactly what work it did. 

```{r testing different joins}
inner_join_drugs <- inner_join(drugs_siezures_tidy,drugs_volume_tidy, by = c("force", "class", "type", "year"))

left_join_drugs <- left_join(drugs_siezures_tidy,drugs_volume_tidy)

right_join_drugs <- right_join(drugs_siezures_tidy,drugs_volume_tidy)

semi_join_drugs <- semi_join(drugs_siezures_tidy,drugs_volume_tidy)

```
