---
title: "Tidyverse_Tibbles"
output: html_document
runtime: shiny
---
# Setup: tidyr #

This .Rmd uses the 'readxl' and 'tidyverse' packages, which I have already loaded for you invisibly in the background. If you are working directly in R (rather than a virtual environment like this) you may need to install and/or load those packages, as well as set your working directory or other start up stuff. 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries and packages, , include=FALSE}
require(readxl)
require(tidyverse)
```

# Loading in the data #

The first step in most R processes is to load in some data. I have provided 4 different data sets in 3 different formats so you can see how to load in different data types.

```{r read in basic style}
drugs_siezures <- read.csv("drugs_siezures.csv",  header = TRUE, check.names=FALSE)
drugs_volume <- read.csv("drugs_volume.csv", header = TRUE, check.names=FALSE)
crime_survey <- read.table("bcs_2007_8_teaching_data.tab", sep="\t", header=TRUE)
sexual_offences <- read_excel("SexualOffences1918_70_Data.xlsx")
```

## Print one of them ##
```{r checking}
print(drugs_siezures)
typeof(drugs_siezures)
```

## Convert all three to tibbles ##
```{r convert to tibbles}
drugs_siezures <- as_tibble(drugs_siezures)
drugs_volume <- as_tibble(drugs_volume)
crime_survey <- as_tibble(crime_survey)
sexual_offences <- as_tibble(sexual_offences)
```

## Print one of them ##
```{r checking again post-tibblization}
print(drugs_siezures)
typeof(drugs_siezures)
```

### Hint = you can do both of these in one step by using this format###
```{r read in advance}
drugs_siezures <- as_tibble(read.csv("drugs_siezures.csv", header = TRUE, check.names=FALSE))
drugs_volume <- as_tibble(read.csv("drugs_volume.csv", header = TRUE, check.names=FALSE))

crime_survey <- as_tibble(read.table("bcs_2007_8_teaching_data.tab", sep="\t", header=TRUE))
sexual_offences <- as_tibble(read_excel("SexualOffences1918_70_Data.xlsx"))
```

# Looking at the data#
## Extract columns of data as values in (at least) 2 different ways ##
```{r examine column data as values}
siezures_drug_type <- drugs_siezures$`Drug type`
typeof(siezures_drug_type)

siezures_drug_class <- drugs_siezures[['Drug class']]
siezures_force <- drugs_siezures[[1]]
typeof(siezures_drug_type)
```

# Looking at the data#
## Extract a column of data as a list in (at least) 2 different ways ##
```{r examine column data as lists}
siezures_2007 <- drugs_siezures[,4]
typeof(siezures_2007)
siezures_2008 <- drugs_siezures[5]
```

## Extract a colum as a variable, extract a row as a variable ##
```{r examine data by row}
Cocaine_siezures <- drugs_siezures[1,]
```