---
title: "Tidyverse_2_Tidyr"
author: "J"
date: "25/01/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(readxl)
require(tidyverse)
options(scipen = 999) 

```

#### Exercises: tidyr ####
First, make sure you have your datasets loaded up as tibbles. 

```{r check loaded as tibbles}
drugs_seizures <- 
  as_tibble(read.csv("drugs_seizures.csv",  header = TRUE, check.names=FALSE, fileEncoding="UTF-8-BOM"))
drugs_volume <- 
  as_tibble(read.csv("drugs_volume.csv", header = TRUE, check.names=FALSE, fileEncoding="UTF-8-BOM"))

crime_survey <- 
  as_tibble(read.table("bcs_2007_8_teaching_data.tab", sep="\t", header=TRUE))
sexual_offences <- 
  as_tibble(read_excel("sexual_offences_1946_1970.xlsx"))
population <- 
  as_tibble(read_excel("population.xlsx"))

```

### Make each tibble tidy as needed ###

```{r gather, separate and relocate drugs}
drugs_seizures_tidy <- drugs_seizures %>% gather(year,seizures,'2007':'2020')
drugs_volume_tidy <- drugs_volume %>% gather(year,volume_seized,'2007':'2020')

drugs_seizures_tidy <- drugs_seizures_tidy %>% separate(drug, c("class","type"), sep='-')
drugs_volume_tidy <- drugs_volume_tidy %>% separate(drug, c("class","type"), sep='-')

drugs_volume_tidy <- drugs_volume_tidy %>% relocate('unit', .after = 'volume_seized')

```

Or do the above in a more streamlined way by piping the results of each step to the next. 

```{r efficiently gather, separate and relocate drugs}
drugs_seizures_tidy <- drugs_seizures %>% gather(year,seizures,'2007':'2020') %>%
  separate(drug, c("class","type"), sep='-')

drugs_volume_tidy <- drugs_volume %>% gather(year,volume_seized,'2007':'2020') %>%
  separate(drug, c("class","type"), sep='-')  %>%
  relocate('unit', .after = 'volume_seized')

```

```{r efficiently gather and separate sexual offences}
sexual_offences_tidy <- sexual_offences %>% 
  gather(offence_name, reports_taken, `abduction-females`:`sex-males`) %>% 
  separate(offence_name, c("offence","against"), sep='-')

```

# The crime_survey data - is it tidy? 
Note: There is no functional code here... You have to look at the crime survey data and decide for yourself whether it needs any gathering, separating or spreading!
```{r gather crime survey}
crime_survey_tidy <- crime_survey %>%
  gather() %>%
  separate()  %>%
  spread()

```

